On the presentation of orthogonal groups over polynomial rings

### 2019-05-04 dl-paper.tex
Added Lemma 13

\subsection{Some elementary lemmas}
\begin{comment}
The following result was proved in \cite[Proposition~3.2]{Ste13}.
\begin{lemma}\label{lem:Zgen} Let $\Phi$ be an irreducible root system of rank $\geq 2$. 
Let $G$ denote the normal closure in $\St(\Phi, R)$ of the subgroup spanned by elementary root unipotents of level $I$, i.\,e. $G=\langle x_\alpha(s) \mid \alpha\in\Phi,\ s\in I\rangle^{\St(\Phi, R)}$.

Let $S$ be a parabolic set of roots of $\Phi$ with special part $\Sigma_S$. Then $G$ is generated as an abstract group by the set $\mathcal{Z}(\Sigma_S, I, R)$ consisting of the following two families of elements:
\begin{itemize}
 \item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
 \item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$.
\end{itemize}        
\end{lemma}
\end{comment}

\begin{comment}
Hall-Witt identity
\[ [[ y^{-1}, x], z] ^ {y^{-1}}  [[ z^{-1}, y], x] ^ {z^{-1}}  [[ x^{-1}, z], y] ^ {x^{-1}} = 1 \]
\[ [[ y^{-1}, x], z]  \cdot [[ z^{-1}, y], x] ^ {z^{-1}y} \cdot  [[ x^{-1}, z], y] ^ {x^{-1}y} = 1. \]
\[ [[ z^{-1}, y^{-1}], x] ^ {z^{-1}y^{-1}} \cdot  [[ x^{-1}, z], y^{-1}] ^ {x^{-1}y^{-1}} = [z, [ y, x]] . \]
\[  . \]
\end{comment}
Throughtout this section $A$ denotes an arbitary ring and $I, J$ two its ideals and $\Phi$ a simply laced root system.

@@

In the following lemma we will need the following variant of Hall--Witt identity:
\begin{equation} \label{HW-variant} [x, [ y, z]] = \left([[ x^{-1}, y^{-1}], z] ^ {x^{-1}} \cdot  [[ z^{-1}, x], y^{-1}] ^ {z^{-1}}\right)^{y^{-1}}.\end{equation}

@@

\subsection{The subgroups $P_\alpha(0)$, $P_\alpha(*)$ and their properties.} Throughout this subsection $\alpha$ denotes some fixed root of $\Phi$.
\begin{df} \label{defP0}
Denote by $P_\alpha(0)$ the subgroup of $\St(\Phi, A[X, X^{-1}])$ generated by the images of the following elements of $\St(\Phi, A[X])$ under the
 map induced by the embedding $A[X] \subseteq A[X, X^{-1}]$:
\begin{enumerate}
 \item $z_{\beta}(Xf, \xi), \beta \in \Phi \text{ such that }\alpha + \beta \in \Phi\text{ or } \alpha - \beta \in \Phi;$
 \item $z_{\beta}(f, X\xi),\ \beta \in \Phi \text{ such that }\alpha - \beta \in \Phi \cup \{0\};$
 \item $z_{\beta}(f, \xi),\ \alpha \perp \beta;$
 \item $x_{-\alpha}(X^2f),$
\end{enumerate}
In the above formulae and below we assume that $f \in M[X]$ and $\xi \in A[X]$. \end{df}

{\bf Warning}: This definition is somewhat incomplete, $P_\alpha(0)$ should also contain certain symbols, e.\,g. those from~\cref{lem:Tulenbayev-formula} below. 
I'll fix this later.

@@

Let us verify the assertion for the generators of type (1) and (2). In the case $\beta - \alpha \in \Phi$ we obtain from Lemma~\ref{Zrels}:
\begin{align} z_{\beta}(Xf, \xi) ^ {x_{-\alpha}(mX)} = & x_{-\alpha} (- mX^2f\xi) \cdot x_{\beta-\alpha} (N_{\beta, -\alpha}\cdot mX^2f) \cdot z_{\beta}(Xf, \xi), \label{eq3-1} \\
  z_{\beta}(f, X\xi) ^ {x_{-\alpha}(mX)} = & x_{-\alpha} (- mX^2f\xi ) \cdot x_{\beta-\alpha} (N_{\beta, -\alpha}\cdot mXf) \cdot z_{\beta}(f, X\xi), \label{eq3-2} \\
  z_{\alpha}(f, X\xi) ^ {x_{-\alpha}(mX)} = & z_{\alpha}(f, X(\xi + m)). \end{align}
The expressions in the right hand side of~\eqref{eq3-1} and~\eqref{eq3-2} are products of generators of type 4, 1, 1 and 4, 1, 2, respectively.

Finally, in the case $\alpha + \beta \in \Phi$ (only possible for the generator of type (1)) we have:
\[ z_{\beta}(Xf, \xi) ^ {x_{-\alpha}(mX)} = x_{-\alpha} (mX^2f\xi ) \cdot x_{-\alpha-\beta} (N_{\beta,\alpha}\cdot mX^2f\xi^2) \cdot z_{\beta}(Xf, \xi). \]
It is clear that the latter expression is a product of generators of type $4, 1, 1$.
\end{proof}

@@

 It remains to notice that in~\eqref{eq:Zconj1}--\eqref{eq:Zconj3} after evaluating at $X=0$ the only terms that remain 
  are those of the form $x_\beta(m)$ for some $\beta \in Z_+(\alpha)$.
\end{proof}

@@

Our next goal is to determine what happens with the generators of $K(A[X], M[X])$ under the map $S_\alpha(a, -)$.
At first we could look at what happens with $K(A[X], M[X]) \cap P_\alpha(0)$ under $S_\alpha(a, -)$.
We notice the following elements:
\begin{itemize}
 \item $x_\alpha(c)$, $c \in A$;
 \item $x_{\alpha+\beta}(d)$, $\alpha+\beta \in \Phi$
\end{itemize}

\[ x_\alpha(c), , x_{\alpha-\beta}(d) \]

>>>>

\subsection{Some elementary lemmas}
\begin{comment}
Hall-Witt identity
\[ [[ y^{-1}, x], z] ^ {y^{-1}}  [[ z^{-1}, y], x] ^ {z^{-1}}  [[ x^{-1}, z], y] ^ {x^{-1}} = 1 \]
\[ [[ y^{-1}, x], z]  \cdot [[ z^{-1}, y], x] ^ {z^{-1}y} \cdot  [[ x^{-1}, z], y] ^ {x^{-1}y} = 1. \]
\[ [[ z^{-1}, y^{-1}], x] ^ {z^{-1}y^{-1}} \cdot  [[ x^{-1}, z], y^{-1}] ^ {x^{-1}y^{-1}} = [z, [ y, x]] . \]
\[  . \]
\end{comment}
Throughtout this section $A$ denotes an arbitary ring and $I, J$ two its ideals and $\Phi$ a simply laced root system.

@@

\begin{comment}
For a subset of roots $S \subseteq \Phi$ denote by $\mathcal{Z}(S, R, I)$ the subset of $\overline{\St}(\Phi, R, I)$ consisting of the following two families of elements:
\begin{itemize}
 \item $x_\alpha(s)$, $\alpha \in \Phi$, $s\in I$;
 \item $z_\alpha(s, \xi)$, $\alpha \in S$, $s\in I$, $\xi \in R$.
\end{itemize}

\begin{lemma} \label{weak-Stepanov} For any $\Phi$ of rank $\geq 2$ the set $\mathcal{Z}(\Phi, R, I)$ generates $\overline{\St}(\Phi, R, I)$ as a group.
 Further, if $\Phi$ is simply laced $\overline{\St}(\Phi, R, I)$ is generated by even smaller set $\mathcal{Z}(\Phi\setminus\{\alpha, \beta\}, R, I)$, 
  where $\alpha, \beta \in \Phi$ are arbitrary roots of $\Phi$. \end{lemma}
\begin{proof}
 The first assertion is well-known, see the first part of~\cite[Lemma~4]{S15}.
 Notice that it follows from relation~\eqref{Z5} that $\mathcal{Z}(S, R, I)$ generates the same group as $\mathcal{Z}(S \cup (S - S), R, I)$.
 The second assertion then follows from the first one and the fact that for $S := \Phi \setminus \{ \alpha, \beta \}$ one has $S \cup (S - S) = \Phi$.
\end{proof}
{\bf TODO:} Probably, in the same fashion we could reprove Stepanov's theorem in the simply laced case.
\end{comment}

\begin{theorem}[Stepanov] \label{thm:Stepanov} For arbitrary $\Phi$ of rank $\geq 2$ the group $\overline{\St}(\Phi, R, I)$ is generated 
by elements $z_\alpha(s, \xi)$, $\alpha \in \Phi$, $s \in I$, $\xi \in R$.

In fact, it is generated by even smaller set of elements defined as follows.
Choose any parabolic subset of roots of $\Phi$ with special part $\Sigma_S$. Then the generating set consists of the following two families of elements:
$x_\alpha(s)$, $s \in I$ and $z_\alpha(s, \xi)$, $\alpha \in \Sigma_S$, $s \in I$, $\xi \in R$. \end{theorem}
\begin{proof} See~\cite[Lemma~4]{S15}. \end{proof}

{\bf TODO:} Perhaps we could deduce the second assertion of Stepanov's theorem from the first one using relation~\eqref{Z5}.

For the proof of the following lemma we will need the following variant of Hall--Witt identity:
\begin{equation} \label{HW-variant} [x, [ y, z]] = \left([[ x^{-1}, y^{-1}], z] ^ {x^{-1}} \cdot  [[ z^{-1}, x], y^{-1}] ^ {z^{-1}}\right)^{y^{-1}}.\end{equation}

@@

\subsection{The subgroups $P_\alpha(0)$, $P_\alpha(*)$ and their properties.} Throughout this subsection $\alpha$ denotes some fixed root of $\Phi$.
\begin{df} \label{defP0}
Denote by $P_\alpha(0)$ the subgroup of $\St(\Phi, A[X, X^{-1}])$ generated by the images of the following elements of $\St(\Phi, A[X])$ under the
 map induced by the embedding $A[X] \subseteq A[X, X^{-1}]$:
\begin{enumerate}
 \item $z_{\beta}(Xf, \xi), \beta \in \Phi \text{ such that }\alpha + \beta \in \Phi\text{ or } \alpha - \beta \in \Phi;$
 \item $z_{\beta}(f, X\xi),\ \beta \in \Phi \text{ such that }\alpha - \beta \in \Phi \cup \{0\};$
 \item $z_{\beta}(f, \xi),\ \alpha \perp \beta;$
 \item $x_{-\alpha}(X^2f),$
\end{enumerate}
In the above formulae and below we assume that $f \in M[X]$ and $\xi \in A[X]$.
{\bf TODO:} the above definition is somewhat incomplete, $P_\alpha(0)$ should also contain certain symbols, e.\,g. those from~\cref{lem:Tulenbayev-formula} below. \end{df}

@@

Let us verify the assertion for the generators of type (1) and (2). In the case $\beta - \alpha \in \Phi$ we obtain from Lemma~\ref{Zrels}:
\begin{align} z_{\beta}(Xf, \xi) ^ {x_{-\alpha}(mX)} = & x_{-\alpha} (- mX^2f\xi) \cdot x_{\beta-\alpha} (N_{\beta, -\alpha}\cdot mX^2f) \cdot z_{\beta}(Xf, \xi), \label{eq3-1} \\
  z_{\beta}(f, X\xi) ^ {x_{-\alpha}(mX)} = & x_{-\alpha} (- mX^2f\xi ) \cdot x_{\beta-\alpha} (N_{\beta, -\alpha}\cdot mXf) \cdot z_{\beta}(f, X\xi), \nonumber \\
  z_{\alpha}(f, X\xi) ^ {x_{-\alpha}(mX)} = & z_{\alpha}(f, X(\xi + m)). \nonumber \end{align}
The expressions in the right hand side of~\eqref{eq3-1} and~\eqref{eq3-2} are products of generators of type 4, 1, 1 and 4, 1, 2, respectively.

Finally, in the case $\alpha + \beta \in \Phi$ (only possible for the generator of type (1)) we have:
\begin{equation} \label{eq3-2} z_{\beta}(Xf, \xi) ^ {x_{-\alpha}(mX)} = x_{-\alpha} (mX^2f\xi ) \cdot x_{-\alpha-\beta} (N_{\beta,\alpha}\cdot mX^2f\xi^2) \cdot z_{\beta}(Xf, \xi). \end{equation}
It is clear that the latter expression is a product of generators of type $4, 1, 1$.
\end{proof}

@@

 It remains to notice that the only factors in the right-hand sides of in~\eqref{eq:Zconj1}--\eqref{eq:Zconj3} which do not vanish after evaluation at $X=0$
  are those of the form $x_\beta(m)$ for some $\beta \in Z_+(\alpha)$.
\end{proof}

In fact, we could prove something a little bit stronger.
\begin{lemma}
 Let $g$ be an element of $K(A[X], M[X])$, then for any pair of roots $\alpha, \beta \in \Phi$, $\alpha \neq \pm \beta$
 the element $g$ can be written as a product $g_0 \cdot x_{-\alpha}(mX) \cdot x_{-\beta}(m'X)$ for some $m, m' \in M$ and
  $g_0 \in K(A[X], M[X]) \cap P_\alpha(0) \cap P_{\beta}(0)$.

 The group $K(A[X], M[X]) \cap P_\alpha(0) \cap P_{\beta}(0)$ admits generating set consisting of the following elements:
 \begin{itemize}
  \item $z_\gamma(Xf, \xi)$, $\gamma \in \Phi \setminus\{ \pm \alpha, \pm \beta\};$
  \item $x_{-\alpha}(X^2f)$, $x_{-\beta}(X^2f)$, $z_{\alpha}(Xf, X\xi)$, $z_\beta(Xf, X\xi)$;
  \item $c_{\pm \delta}(f, X\xi)$ for some fixed root $\delta \in \Phi$, $\delta \not \in \{ \pm \alpha, \pm \beta \}$
 \end{itemize}
\end{lemma}
\begin{proof}
 %Follows from~\cref{Kgen-strong}, \cref{P0_normal} and Stepanov's theorem.
 Embed $\alpha$ and $\beta$ in some subsystem $\Psi$ of type $A_3$ and denote by $\Sigma$ the special part of the parabolic subset $S = \Psi \cup \Phi^+$.
 Applying~\cref{thm:Stepanov} to $S$ we obtain that every element of $\St(\Phi, A[X], XM[X])$ can be presented as a product of 
  $z_\gamma(Xf, \xi)$, $\gamma \in \Sigma(S) \subseteq \Phi^+\setminus \{ \pm \alpha, \pm \beta \}$ and $x_\gamma(Xf)$, $\gamma \in \{ \pm \alpha, \pm \beta \}$.

 Now choose arbitrary $\delta \neq \pm \alpha, \pm \beta$. By~\cref{Kgen-strong} we can write $g$ as a product of elements of the above family and
  generators of the form $c_{\pm \delta}(f, X\xi)$.

 The next step of the proof is to write each factor of the form $x_{\epsilon}(Xf)$, $\epsilon \in \{-\alpha,-\beta\}$ appearing in this product as
 $x_{\epsilon}(X^2f') \cdot x_{\epsilon}(mX)$ and then move $x_{\epsilon}(mX)$ to the rightmost position
  (conjugating by $x_{\epsilon}(mX)$ all factors along the way).
 Notice that no terms of the form $x_{\epsilon}(mX)$, $\epsilon \in \{ -\alpha, -\beta \}$ may appear after the simplification of conjugates.
 Indeed, notice that the extra terms in the right-hand sides of \eqref{eq3-1} and~\eqref{eq3-2} 
 (as well as all terms in the right-hand sides of~\eqref{C1},\eqref{C2} of~\cref{Crels} if you substitute proper values into it)
  have terms of degree at least $2$ in $X$. As a side effect of this operation, however, terms $x_\alpha(Xf)$ and $x_\beta(Xf)$ appearing in $g$ could have turned into
   $z_\alpha(Xf, X\xi)$, and $z_\beta(Xf, X\xi)$ for some $\xi \in A[X]$.
 After collecting terms of the form $x_\epsilon(mX)$, $\epsilon \in \{ -\alpha, -\beta \}$ in the right-hand side of the resulting expression
  we obtain the desired decomposition for $g$. \end{proof}

@@
